Phylodynamic inference of contact network parameters through approximate Bayesian computation

### 2016-06-07 doc/thesis/discussion.Rnw

@@

With respect to the preferential attachment power $\alpha$, the six datasets
analysed fell into two categories (\cref{fig:abchpd}). First, we estimated a
preferential attachment power close to 1, indicating linear preferential
attachment, for the BC data and the outbreaks studied by
\textcite{niculescu2015recent} and \textcite{wang2015targeting}. These values
were robust to specifying different priors for $m$. All three datasets were
sampled from populations in which we would expect a high degree of
epidemiological relatedness: \textcite{niculescu2015recent} studied a recent
outbreak among Romanian \gls{IDU}, \citeauthor{wang2015targeting} sampled
acutely infected MSM in Beijing, China, and the BC data constituted a
phylogenetic \gls{IDU} cluster. These are all contexts in which we would expect
some of the assumptions of the BA model, such as a connected network,
relatively high mean degree, and preferential attachment dynamics, to hold.

The remaining three datasets (\textcite{cuevas2009hiv, novitsky2014impact,
li2015hiv}) had estimated values of $\alpha$ below 0.5 when $m = 1$ was
included in the prior, but these were not robust to changing the prior to
exclude $m = 1$. For the \citeauthor{cuevas2009hiv} data, model
misspecification is likely partially responsible. While the authors found that
a large proportion of the samples were epidemiologically linked, these were
mainly in small local clusters rather than the single large component
postulated by the BA model. In addition, the mixed risk groups in the dataset
would be unlikely to significantly interact, further weakening any global
preferential attachment dynamics. The dataset studied by
\textcite{novitsky2014impact} originated from a densely sampled population
where the predominant risk factor was believed to be heterosexual exposure.
Although the MAP estimate of $\alpha$ was almost unchanged when the value $m =
1$ was excluded from the prior, the confidence interval shrank substantially.

{\color{blue}\uline{
The posterior mean estimates for the real world \gls{HIV} datasets corresponded
to power law exponents $\gamma$ between 2.09 and 2.49.  TODO
}}

>>>>

<<setup, include=FALSE>>=
    source("global.R")
@

@@

{\color{red}\sout{With respect to the preferential attachment power $\alpha$,
the six datasets analysed fell into two categories} (\cref{fig:abchpd}).
\sout{First, we estimated a preferential attachment power close to 1,
indicating linear preferential attachment, for the BC data and the outbreaks
studied by} \textcite{niculescu2015recent} and \textcite{wang2015targeting}.
\sout{These values were robust to specifying different priors for $m$. All three
datasets were sampled from populations in which we would expect a high degree
of epidemiological relatedness:} \textcite{niculescu2015recent} \sout{studied a
recent outbreak among Romanian \gls{IDU},} \citeauthor{wang2015targeting}
\sout{sampled acutely infected MSM in Beijing, China, and the BC data
constituted a phylogenetic \gls{IDU} cluster. These are all contexts in which
we would expect some of the assumptions of the BA model, such as a connected
network, relatively high mean degree, and preferential attachment dynamics, to
hold.}

\sout{The remaining three datasets} (\textcite{cuevas2009hiv,
novitsky2014impact, li2015hiv}) \sout{had estimated values of $\alpha$ below
0.5 when $m = 1$ was included in the prior, but these were not robust to
changing the prior to exclude $m = 1$. For the} \citeauthor{cuevas2009hiv} 
\sout{data, model misspecification is likely partially responsible. While the
authors found that a large proportion of the samples were epidemiologically
linked, these were mainly in small local clusters rather than the single large
component postulated by the BA model. In addition, the mixed risk groups in the
dataset would be unlikely to significantly interact, further weakening any
global preferential attachment dynamics. The dataset studied by}
\textcite{novitsky2014impact} \sout{originated from a densely sampled
population where the predominant risk factor was believed to be heterosexual
exposure. Although the MAP estimate of $\alpha$ was almost unchanged when the
value $m = 1$ was excluded from the prior, the confidence interval shrank
substantially.}}

<<disc, include=FALSE>>=
    d <- setDT(collect.data("../../simulations/aggregates/hpd/*"))
@

For all datasets we examined, the posterior mean estimates for \gls{alpha} were
sub-linear, ranging from 
    \Sexpr{d[m_min == 1 & parameter == "alpha", round(min(mean), 2)]}
to 
    \Sexpr{d[m_min == 1 & parameter == "alpha", round(max(mean), 2)]}.
The sub-linearity is consistent with the results of
\textcite{de2007preferential}, who developed a statistical inference method to
estimate the parameters of a more sophisticated \acrlong{PA} model
incorporating heterogeneous node behaviour. They found \gls{alpha} values
ranging from 0.26 to 0.62, depending on the gender and time period considered. 
Our estimates of \gls{alpha} for the \textcite{niculescu2015recent} was above
this range under both priors, as were the estimates for the
\textcite{wang2015targeting} data and the BC data when \gls{m} = 1 was
disallowed by the prior. The dataset investigated by
\textcite{de2007preferential} was derived from a survey of a random sample of
the Norwegian population, whereas our investigation focused on datasets from
known phylogenetic or geographic clusters of \gls{HIV} infected persons. It is
therefore unsurprising that we detected stronger \acrlong{PA} dynamics in some
cases. For instance, random sampling is much less likely to discover the
high-degree nodes characterizing the tail of the degree distribution, simply
because those individuals are rare in the general population. In addition, it
is plausible that \gls{HIV}-positive individuals are more likely to be highly
connected in their sexual networks, as the odds of acquiring \gls{HIV} increase
with the number of unprotected sexual contacts.

Both \textcite{de2007preferential} and \textcite{novitsky2014impact} studied
populations whose primary risk factor for \gls{HIV} infection was heterosexual
contact. \citeauthor{de2007preferential} explicitly excluded reported
homosexual contacts; \citeauthor{novitsky2014impact} did not, but noted that
heterosexual contact is the primary mode of transmission in Botswana where the
study was done. In the first of the two papers describing the Botswana
study~\autocite{novitsky2013phylogenetic}, the authors noted that their sample
was gender-biased, being composed of approximately 75\% women. Our estimate of
\gls{alpha} for these data was X or X, depending on the prior on \gls{m};
\citeauthor{de2007preferential} estimated 0.54, 0.57, and 0.29 for 3-year,
5-year, and lifetime partnership networks respectively for the female portion
of their sample.

For both choices of prior on \gls{m}, the datasets derived from \gls{IDU}
populations had a higher estimated \acrlong{PA} power than the other datasets
(\cref{fig:abchpd,fig:abchpdm2}). This finding is in line with
\textcite{dombrowski2013topological}, who reanalyzed a network of \glspl{IDU}
in Brooklyn, USA, collected between 1991 and
1993~\autocite{friedman2006social}. They found that the the \gls{IDU} network 
resembled a \gls{BA} network much more closely than other social and sexual 
networks, and offered sociological explanations for the apparent \acrlong{PA}
dynamics in this population. Importantly, from a public health perspective,
the authors asserted that the removal of \emph{random} individuals from
\gls{IDU} networks may have the paradoxical effect of increasing the network's
epidemic susceptibility. When low-degree nodes are removed, as would occur
during a police crackdown, their network neighbours may turn to well-known
community members for advice or supplies, thus increasing the connectivity of
these high-degree nodes.
